---
title: "StatisticalAnalysis"
output: html_document
---

```{r}
# setup
library(tidyverse)
library(broom)
library(ggplot2)
install.packages("betareg")
library(betareg)
library(broom)
install.packages("emmeans")
library(emmeans)
```

# Beta Regression

```{r}
getwd()
df <- read.csv("../data/final_df_sentiment.csv")
```

```{r}
df_long <- df %>% 
  select(ID, prob_pos_context, prob_pos_human_response, prob_pos_ft_response, prob_pos_gpt_response
  ) %>% 
  pivot_longer(
    cols = c(prob_pos_human_response, prob_pos_ft_response, prob_pos_gpt_response),
    names_to = "response_type",
    values_to = "response_prob"
  ) %>%
  mutate(
    context_prob = prob_pos_context,
    is_ft = ifelse(response_type == "prob_pos_ft_response", 1, 0),
    is_gpt = ifelse(response_type == "prob_pos_gpt_response", 1, 0)
  )
```

```{r}
# fix any probabilities that are 0 or 1 - need to be within 0-1
eps <- 1e-5
df_long$response_prob <- (df_long$response_prob * (1 - 2*eps)) + eps
df_long$context_prob  <- (df_long$context_prob * (1 - 2*eps)) + eps

```


### Models
```{r}
m3 <- betareg(
  response_prob ~ context_prob + is_ft + is_gpt,
  data = df_long
)

summary(m3)


m4 <- betareg(
  response_prob ~ context_prob * is_ft + context_prob * is_gpt,
  data = df_long
)

summary(m4)

```
INTERPRETATION GUIDE

Coefficients are on a log-odds scale (like logistic)
Positive coefficients → higher predicted positivity
Negative coefficients → lower predicted positivity.

Key terms:
context_prob     : more positive context → more positive response
is_ft            : FT more (or less) positive than human
is_gpt           : GPT more (or less) positive than human
context_prob:is_ft   : FT differs from humans in coherence strength
context_prob:is_gpt  : GPT differs from humans in coherence strength


```{r}
# Test/compare models 
lrtest(m3, m4)  # Likelihood ratio test
   AIC(m3, m4)     # Compare model fit
```

```{r}
# Check model assumptions
plot(m3)  # Residual diagnostics
```

```{r}
# Predicted Probabilities - compare responses under negative context (0) and positive context (1)
# Create prediction data - one row per combination
pred_data <- data.frame(
  context_prob = rep(c(0.01, 0.99), each = 3),
  is_ft = rep(c(0, 1, 0), times = 2),
  is_gpt = rep(c(0, 0, 1), times = 2),
  response_type = rep(c("Human", "FT", "GPT"), times = 2)
)

# Get predictions
pred_data$predicted_prob <- predict(m3, newdata = pred_data, type = "response")

print(pred_data)

```

```{r}
# Visualise

ggplot(pred_data, aes(
  x = response_type,
  y = predicted_prob,
  fill = as.factor(context_prob)
)) +
  geom_col(position = position_dodge()) +
  labs(
    x = "Responder Type",
    y = "Predicted Positivity Probability",
    fill = "Context Sentiment\n(0 = Negative, 1 = Positive)",
    title = "Predicted Sentiment Probabilities from Beta Regression"
  ) +
  ylim(0,1)

```


### Post-hoc contrasts
Compare predicted probabilities between groups (FT vs Human, GPT vs Human) for a given context.
```{r}
# separate diff sentiment contexts
# Create prediction data - one row per combination
pred_data <- data.frame(
  context_prob = rep(c(0.01, 0.99), each = 3),
  is_ft = rep(c(0, 1, 0), times = 2),
  is_gpt = rep(c(0, 0, 1), times = 2),
  response_type = rep(c("Human", "FT", "GPT"), times = 2)
)

# Get predictions
pred_data$predicted_prob <- predict(m3, newdata = pred_data, type = "response")
```

### Get contrasts
```{r}
# Separate by context
pred_neg <- pred_data %>% filter(context_prob < 0.5)
pred_pos <- pred_data %>% filter(context_prob > 0.5)

# Compute point estimate contrasts
contrasts <- data.frame(
  context = c("Negative", "Negative", "Positive", "Positive"),
  contrast = c("FT_vs_Human", "GPT_vs_Human", "FT_vs_Human", "GPT_vs_Human"),
  difference = c(
    pred_neg$predicted_prob[pred_neg$response_type == "FT"] - 
      pred_neg$predicted_prob[pred_neg$response_type == "Human"],
    pred_neg$predicted_prob[pred_neg$response_type == "GPT"] - 
      pred_neg$predicted_prob[pred_neg$response_type == "Human"],
    pred_pos$predicted_prob[pred_pos$response_type == "FT"] - 
      pred_pos$predicted_prob[pred_pos$response_type == "Human"],
    pred_pos$predicted_prob[pred_pos$response_type == "GPT"] - 
      pred_pos$predicted_prob[pred_pos$response_type == "Human"]
  )
)

print(contrasts)
```


```{r}
# Visualize
ggplot(contrasts, aes(x = contrast, y = difference, fill = context)) +
  geom_col(position = position_dodge(0.9)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Contrast",
    y = "Difference in predicted probability",
    fill = "Context",
    title = "AI vs Human Response Sentiment"
  ) +
  theme_minimal()
```

Interpretation:
If there is a positive difference = FT or GPT more positive than Human
If there is a negative difference = FT or GPT more negative than Human
Dashed line at 0 is the Human baseline
Shows how much each AI differs from humans under negative vs positive context.
